The Sloan Digital Sky Survey Spectroscopic Reduction Pipeline

### 2006-03-31 doc/paper/paper2d.tex
Update.

The extracted science spectra are first normalized by the fiber flat vectors.
Strong atmospheric emission lines (sky lines) are identified in the the 
set of spectra, and the wavelength solution is shifted and stretched 
to match the identified sky line positions.  
Examples of the shift corrections are shown in Figure ??.  
The widths of the identified sky lines are also measured with the same method
as the arc-line width measurements.  All spectra are also normalized by
the average flat-field spectrum (the superflat) to remove small scale
features likely attributed to the chromatic response of the dichroic.
The wavelength solution is corrected to both vacuum wavelengths and the
heliocentric velocity frame.

@@

@@

@@

%------------------------------------------------------------------------------
%------------------------------------------------------------------------------
\appendix
\section{DATA MODEL}
\label{bsplines}

What the heck is a B-spline?

>>>>

The extracted science spectra are first normalized by the fiber flat vectors.
Strong atmospheric emission lines (sky lines) are identified in the the 
set of spectra, and the wavelength solution is shifted and stretched 
to match the identified sky line positions.  
Examples of the shift corrections are shown in Figure ??.  
The widths of the identified sky lines are also measured with the same method
as the arc-line width measurements.  All spectra are also normalized by
the average flat-field spectrum (the superflat) to remove small scale
features likely attributed to the chromatic response of the dichroic.
The wavelength solution is corrected to both vacuum wavelengths and the
barycenter velocity frame.

@@

\section{Testing velocities}

% See sdss-spcommish/206 for description of plate 321

A plate was drilled centered on the open cluster M67 with the
specific idea of testing the SDSS velocity precision.
130 stars were targetted from the radial velocity catalog
of Mathieu \etal\ (1986), falling in the magnitude range of 8 to 14.
The 19 stars brighter than $m=11$ were not plugged, since they
would have saturated the detectors.  This was plate 321 MJD 51612.

@@

% Precise radial velocities of late-type stars in the open clusters M11 and M67
\bibitem[Mathieu \etal\ (1896)]{1986AJ...92...1100} Mathieu, R.~D.,
Latham, D.~W., Griffin, R.~F., \& Gunn, J.~E.\ 1986, \aj, 92, 1100

@@

%------------------------------------------------------------------------------
%------------------------------------------------------------------------------
\section{DATA MODEL}
\label{datamodel}

This spCFrame files contain the extracted spectra for a single CCD of a
single exposure. In general, each object has several blue exposures and
several red exposures. These spectra are still in the native wavelength
mapping, which is neither linear in wavelength nor log-wavelength.
However, the spectra are flux-calibrated, and deviant pixels are
rejected in the pixel mask (from the comparison with the other exposures).
The spCFrame file contains the following:
\begin{verbatim}
HDU #0 = Flux in units of 10^(-17) erg/s/cm^2/Ang [FLOAT]
HDU #1 = Inverse variance (1/sigma^2) for the above [FLOAT]
HDU #2 = Pixel mask [32-bit INT]
HDU #3 = Wavelength solution in log-10 Angstroms (vacuum) [FLOAT]
HDU #4 = Wavelength dispersion in units of (10e-4 log-10 wavelength)^(-1) [FLOAT]
         (these are the same units as in the combined-spectrum spPlate file)
HDU #5 = Plug-map structure from plPlugMapM file [BINARY FITS TABLE]
HDU #6 = Sky flux in units of 10^(-17) erg/s/cm^2/Ang [FLOAT]
\end{verbatim}

The spPlate files contains the combined spectra for all exposures
(potentially spanning multiple nights) for a given mapped plate.
The first 5 HDU's contain NPIX by 640 images, where NPIX is the number
of wavelengths. The pixel scale is exactly 1e-4 in log-wavelength
(any units), which corresponds to $10^{-4} c \ln(10) = 69.029765 \kms$.
The spPlate file contains the following:
\begin{verbatim}
HDU #0 = Flux in units of 10^(-17) erg/s/cm^2/Ang [FLOAT]
HDU #1 = Inverse variance (1/sigma^2) for the above [FLOAT]
HDU #2 = AND mask [32-bit INT]
HDU #3 = OR mask [32-fit INT]
HDU #4 = Wavelength dispersion in pixels [FLOAT]
HDU #5 = Plug-map structure from plPlugMapM file [BINARY FITS TABLE]
HDU #6 = Average sky flux in units of 10^(-17) erg/s/cm^2/Ang [FLOAT]
\end{verbatim}

The following header keywords are of interest (in HDU \#0):
\begin{verbatim}
PLATEID = Plate ID
TILEID  = Tile ID
MJD     = Modified Julian date for first night of observation
MJDLIST = Modified Julian date for all nights of observation
MAPID   = Fiber mapper ID
CARTID  = Fiber cartridge used at the telescope
NAME    = Name of target: Plate ID, MJD, Fiber mapper rerun
COEFF0  = Central wavelength (log10-Angstroms) of first pixel
COEFF1  = Wavelength range of each pixel (log-10 Angstroms)
CRVAL1  = (Same as COEFF0)
CD1_1   = (Same as COEFF1)
RADEG   = Right ascension of plate center (J2000 degrees)
DECDEG  = Declination of plate center (J2000 degrees)
AIRMASS = Airmass for mid-point of observation
RA      = Right ascension of where the telescope thinks it is pointing
DEC     = Declination of where the telescope thinks it is pointing
AZ      = Azimuth of the telescope (degrees)
ALT     = Altitude of the telescope (degrees)
NEXP    = Total number of frames from all cameras used in the co-adding;
          since there are 4 cameras, NEXP=12 would typically mean that there
          were 3 exposures in each camera.
EXPT_B1 = b1 camera exposure time (seconds)
EXPT_B2 = b2 camera exposure time (seconds)
EXPT_R1 = r1 camera exposure time (seconds)
EXPT_R2 = r2 camera exposure time (seconds)
EXPTIME = Minimum of exposure times for all cameras
SPEC1_G = (S/N)^2 per pixel at g=20.2 on spectrograph #1
SPEC2_G = (S/N)^2 per pixel at g=20.2 on spectrograph #2
SPEC1_R = (S/N)^2 per pixel at r=20.25 on spectrograph #1
SPEC2_R = (S/N)^2 per pixel at r=20.25 on spectrograph #2
SPEC1_I = (S/N)^2 per pixel at i=19.9 on spectrograph #1
SPEC2_I = (S/N)^2 per pixel at i=19.9 on spectrograph #2
GOFFSTD = Median spectrophotometry minus photo mag for standard stars in g-band
ROFFSTD = Median spectrophotometry minus photo mag for standard stars in r-band
IOFFSTD = Median spectrophotometry minus photo mag for standard stars in i-band
GOFFRMS = RMS of spectrophotometry minus photo mag for standard stars in g-band
ROFFRMS = RMS of spectrophotometry minus photo mag for standard stars in r-band
IOFFRMS = RMS of spectrophotometry minus photo mag for standard stars in i-band
GOFFGAL = Median spectrophotometry minus photo mag for "main" galaxies in g-band
ROFFGAL = Median spectrophotometry minus photo mag for "main" galaxies in r-band
IOFFGAL = Median spectrophotometry minus photo mag for "main" galaxies in i-band
GOFFGAL = RMS of spectrophotometry minus photo mag for "main" galaxies in g-band
ROFFGAL = RMS of spectrophotometry minus photo mag for "main" galaxies in r-band
IOFFGAL = RMS of spectrophotometry minus photo mag for "main" galaxies in i-band
BESTEXP = Exposure number for the "best" exposure (all 4 cameras)
EXPIDxx = Identifier for the each invididual exposure+CCD used in this combine,
          for ex. 'b1-00006853-00006857-00006858' refers to the b1 CCD,
          exposure #6853, using flat #6857 and arc #6858
TAI     = Number of seconds since Nov 17 1858 for mid-point of observation
TAI-BEG = Number of seconds since Nov 17 1858 for beginning of observation
TAI-END = Number of seconds since Nov 17 1858 for end of observation
SEEING20= 20th-percentile of seeing during exposure (arcsec)
SEEING50= 50th-percentile of seeing during exposure (arcsec)
SEEING80= 80th-percentile of seeing during exposure (arcsec)
RMSOFF20= 20th-percentile of RMS offset of guide fibers (arcsec)
RMSOFF50= 50th-percentile of RMS offset of guide fibers (arcsec)
RMSOFF80= 80th-percentile of RMS offset of guide fibers (arcsec)
VERSIDL = Version of IDL
VERS2D  = Version of idlspec2d for 2D reduction
VERSCOMB= Version of idlspec2d for combining multiple spectra
VERSFLAT= Version of specflat for spectroscopic flat-fields
VERS1D  = Version of idlspec2d for 1D reduction
HELIO_RV= Barycentric velocity correction (km/sec)
\end{verbatim}

There are two masks, an "AND" mask and an "OR" mask. The spectra
are constructed from 3 or more 15-minute observations, and the
"AND" mask bits are set if that bit is set for each and every
input observation. The "OR" mask bits are set if that bit is set
for any of the observations. Usually, I only look at the "AND" mask.
The mask bits are set as follows:
\begin{verbatim}
 0 NOPLUG          Fiber not listed in plugmap file
 1 BADTRACE        Bad trace from routine TRACE320CRUDE
 2 BADFLAT         Low counts in fiberflat
 3 BADARC          Bad arc solution
 4 MANYBADCOLUMNS  >10% pixels are bad columns
 5 MANYREJECTED    >10% pixels are rejected in extraction
 6 LARGESHIFT      Large spatial shift between flat and object pos'n
 7 BADSKYFIBER     Sky Fiber shows extreme residuals
 8 NEARWHOPPER     Within 2 fibers of a whopping fiber (inclusive)
 9 WHOPPER         Whopping fiber
16 NEARBADPIXEL    Bad pixel within 3 pixels of trace
17 LOWFLAT         Flat field less than 0.5
18 FULLREJECT      Pixel fully rejected in extraction (INVVAR=0)
19 PARTIALREJECT   Some pixels rejected in extraction
20 SCATTEREDLIGHT  Scattered light significant
21 CROSSTALK       Cross-talk significant
22 NOSKY           Sky level unknown at this wavelength (INVVAR=0)
23 BRIGHTSKY       Sky level > flux + 10*(flux error)
                   AND sky > 2.0 * median(sky,99 pixels)
24 NODATA          No data available in combine B-spline (INVVAR=0)
25 COMBINEREJ      Rejected in combine B-spline
26 BADFLUXFACTOR   Low flux-calibration or flux-correction factor
27 BADSKYCHI       Relative chi^2 > 3 in sky residuals at this wavelength
28 REDMONSTER      Contiguous region of bad chi^2 in sky residuals
                   (with threshhold of relative chi^2 > 3)
\end{verbatim}
When low numbered bits ($<16$) are set, those will be set for half of the
spectra -- either the blue or red spectrograph. The higher-numbered bits
($>=16$) are set for individual pixels.

Which mask bits are important?
The conditions that are considered very bad are already used to set
the errors to infinity for the effected pixels (specifically, the
inverse variance is set to zero). The most useful mask bit to look
at is BRIGHTSKY, which indicates when the sky is so bright relative
to the object that perhaps one shouldn't trust any of the object
flux there. Our reported errors are meant to include sky-subtraction
errors, but there are instances (particularly around 5577) where
these errors may be untrustworthy.

The dispersion per pixel and the sky flux are computed at each pixel
by re-weighting the individual spectra at each pixel according to
their formal errors. This re-weighting is only approximate.

Note that the sky lines are slightly shifted in the reductions
because we transform the velocities to the barycenter of the solar
system. Each exposure that contributes to the co-added spectra will
have slightly different barycenter correction, so the "average sky"
contains a superposition of these slightly-offset sky lines. These
shifts keep the object spectra as-measured at the barycenter,
regardless of the time of year or the Earth's rotation relative to
the spectroscopic targets.

\subsection{Target Flags}

The target selection sets bits in two flags called PRIMTARGET and
SECTARGET. An object can have multiple bits set. For example, if
the target selection code determined that an object was both a
QSO\_SKIRT and a STAR\_WHITE\_DWARF, then two bits will be set. The
value of PRIMTARGET in this example will be $2^4 + 2^{19} = 524304$.

The PRIMTARGET bit values are as follows (in hex and decimal):
\begin{verbatim}
  QSO_HIZ           = 0x1          = bit # 0 =        1
  QSO_CAP           = 0x2          = bit # 1 =        2
  QSO_SKIRT         = 0x4          = bit # 2 =        4
  QSO_FIRST_CAP     = 0x8          = bit # 3 =        8
  QSO_FIRST_SKIRT   = 0x10         = bit # 4 =       16
  QSO_MAG_OUTLIER   = 0x2000000    = bit #25 = 33554432
  QSO_REJECT        = 0x20000000   = bit #29 =536870912
  GALAXY_RED        = 0x20         = bit # 5 =       32
  GALAXY_RED_II     = 0x4000000    = bit #26 = 67108864
  GALAXY            = 0x40         = bit # 6 =       64
  GALAXY_BIG        = 0x80         = bit # 7 =      128
  GALAXY_BRIGHT_CORE= 0x100        = bit # 8 =      256
  ROSAT_A           = 0x200        = bit # 9 =      512
  ROSAT_B           = 0x400        = bit #10 =     1024
  ROSAT_C           = 0x800        = bit #11 =     2048
  ROSAT_D           = 0x1000       = bit #12 =     4096
  ROSAT_E           = 0x8000000    = bit #27 =134217728
  STAR_BHB          = 0x2000       = bit #13 =     8192
  STAR_CARBON       = 0x4000       = bit #14 =    16384
  STAR_BROWN_DWARF  = 0x8000       = bit #15 =    32768
  STAR_SUB_DWARF    = 0x10000      = bit #16 =    65536
  STAR_CATY_VAR     = 0x20000      = bit #17 =   131072
  STAR_RED_DWARF    = 0x40000      = bit #18 =   262144
  STAR_WHITE_DWARF  = 0x80000      = bit #19 =   524288
  STAR_PN           = 0x10000000   = bit #28 =268435456
  SERENDIP_BLUE     = 0x100000     = bit #20 =  1048576
  SERENDIP_FIRST    = 0x200000     = bit #21 =  2097152
  SERENDIP_RED      = 0x400000     = bit #22 =  4194304
  SERENDIP_DISTANT  = 0x800000     = bit #23 =  8388608
  SERENDIP_MANUAL   = 0x100000     = bit #24 = 16777216
  QSO_MAG_OUTLIER   = 0x200000     = bit #25 = 33554432
  GALAXY_RED_II     = 0x400000     = bit #26 = 67108864
  ROSAT_E           = 0x800000     = bit #27 =134217728
  STAR_PN           = 0x1000000    = bit #28 =268435456
  QSO_REJECT        = 0x2000000    = bit #29 =536870912
  SOUTHERN_SURVEY   = 0x8000000    = bit #31 =-2147483648
\end{verbatim}

The SECTARGET bit values are as follows (in hex and decimal):
\begin{verbatim}
  LIGHT_TRAP        = 0x1          = bit # 0 =        1
  REDDEN_STD        = 0x2          = bit # 1 =        2
  TEST_TARGET       = 0x4          = bit # 2 =        4
  QA                = 0x8          = bit # 3 =        8
  SKY               = 0x10         = bit # 4 =       16
  SPECTROPHOTO_STD  = 0x20         = bit # 5 =       32
  GUIDE_STAR        = 0x40         = bit # 6 =       64
  BUNDLE_HOLE       = 0x80         = bit # 7 =      128
  QUALITY_HOLE      = 0x100        = bit # 8 =      256
  HOT_STD           = 0x200        = bit # 9 =      512
  SOUTHERN_SURVEY   = 0x8000000    = bit #31 =-2147483648
\end{verbatim}

\subsection{Vacuum wavelengths}

The data are stored using vacuum wavelengths.
Most optical astronomers know the wavelengths of transitions
as measured at S.T.P., which is how the CRC lists them for any
transitions redward of $2000\AA$.

The IAU standard for conversion from air to vacuum wavelengths
is given in Morton (1991, ApJS, 77, 119). For vacuum wavelengths ($\lambda$)
in Angstroms, convert to air wavelength ($\lambda_{\rm air}$) via:
$$\lambda_{\rm air} = \lambda / (1 + 2.735182E-4 + 131.4182 / \lambda^2 + 2.76249E8 / \lambda^4) $$

Below, I've listed the vacuum wavelength of some transitions:
\begin{verbatim}
 Air       Vacuum
 --------  --------
 4861.363  4862.721  H_beta
 4958.911  4960.295  [O III]
 5006.843  5008.239  [O III]
 6548.05   6549.86   [N II]
 6562.801  6564.614  H_alpha
 6583.45   6585.27   [N II]
 6716.44   6718.29   [S II]
 6730.82   6732.68   [S II]
\end{verbatim}
The wavelengths are shifted such that measured velocities are relative
to barycentric at the mid-point of each 15-minute exposure.
